Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2017-07-31 manuscript/manuscript.tex
added plots

\begin{abstract}
Supermassive black hole binaries (SMBHBs) are predicted to form a high-redshift cosmological population of low-frequency gravitational wave sources which added together produce a stochastic gravitational wave background (GWB) potentially measurable via pulsar timing. Comparing the actual and predicted times-of-arrival (TOAs) of radio pulses beamed by millisecond pulsars allows us to search for and place limits on the stochastic background. While most estimating of GWB significance is done via Bayesian analysis, here we use a frequentist estimator called the noise-marginalized optimal statistic to assess the signifcance of a GWB present in simulated PTA data. We find \ki{complete}

@@

Pulsar timing arrays (PTAs) are a conglomerate of pulsars with millisecond periods. Monitioring the ultra-regular emission from these pulsars with radio telescopes enables us to probe the dynamics of the spacetime through which the radio pulses travel. This makes PTAs ideal complements to ground- and space-based gravitational wave (GW) detectors. A PTAs sensitivity is constrained by single pulsar observation cadence and total experiment length. For NANOGrav, this means GW frequencies in the range $10^{-9} - 10^{-7}$ Hz. \ki{citations needed} predict that the incoherent superposition of inspiraling SMBHBs form the dominant signal in this low-frequency regime.

@@

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background, 
and Fig.~\ref{fig:os_datasetstats} compares the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for each realization of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 7.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.2 \times 10^{-29}$.

@@

\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for \sv{insert number} simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 7.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.2 \times 10^{-29}$.}
	\label{fig:os_datasetstats}
\end{figure}

@@

Figure~\ref{fig:} shows the mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and Hellings-Downs spatial correlations 
for \sv{insert number} simulated data sets.

>>>>

\begin{abstract}
Supermassive black hole binaries (SMBHBs) are predicted to form a high-redshift cosmological population of low-frequency gravitational wave sources which added together produce a stochastic gravitational wave background (GWB) potentially measurable via pulsar timing. Comparing the actual and predicted times-of-arrival (TOAs) of radio pulses beamed by millisecond pulsars allows us to search for and place limits on the stochastic background. While most estimating of GWB significance is done via Bayesian analysis, here we use a frequentist estimator called the noise-marginalized optimal statistic to assess the significance of a GWB present in simulated PTA data. We find \ki{complete}

@@

Pulsar timing arrays (PTAs) are a conglomerate of pulsars with millisecond periods. 
Monitoring the ultra-regular periodic emission from these pulsars with radio telescopes 
enables us to probe the dynamics of the spacetime through which the pulses travel. 
PTAs are sensitive to gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7}\;\mathrm{Hz}$ \ki{citations needed}, 
which makes PTAs ideal complements to ground- and space-based GW detectors \sv{add citations}. 
\sv{something...} predict that the incoherent superposition of inspiraling SMBHBs form the dominant signal in this low-frequency regime.

@@

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background, 
and Fig.~\ref{fig:os_datasetstats} compares the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for each realization of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$.

@@

\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_datasetstats}
\end{figure}

@@

Figure~\ref{fig:os_ORF} shows the mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and Hellings-Downs spatial correlations 
for 425 simulated data sets.

\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_datasetstats_5e-15_ORF.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), and quadrupole (green) 
			spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}
